Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2019-12-18 article-information-2019.tex
many edits/improvements

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using adverse impact ratio (AIR), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: AIR, SMD, and other group fairness metrics -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository associated with Sections \ref{sec:m_and_m} and \ref{sec:res} -- \textsection\ref{ssec:soft}
\end{itemize}

To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis.

Post-hoc explanation and discrimination testing techniques are applied to constrained, interpretable models trained on the mortgage data to provide a more realistic template workflow for future users of similar methods and tools. Presented explanation techniques include PD, ICE, and Shapley values \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods include measures of DI with marginal effects, AIR and SMD \cite{feldman2015certifying}, \cite{cohen2013statistical}, \cite{cohen1992power}.\footnote{Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix metrics are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available in the software resources associated delineated in Subsection \ref{ssec:soft}.

@@

@@

The mortgage dataset analyzed here is a random sample of 100,000 consumer-anonymized loans from the HDMA database, separated into a training set and a test set. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e. the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider.  Ultimately, the HMDA data represent the most comprehensive source of data on highly-regulated mortgage lending that is publicly available, which makes it an ideal dataset to use for the types of analyses set forth in this text.

@@

\begin{equation}
\begin{aligned}\label{eq:rf}
g^{\text{MGBM}}(\mathbf{x}) &= \sum_{b=1}^B T_b\left(\mathbf{x}; \Theta^\text{mono}_b\right)
\end{aligned}
\end{equation}

@@

Following \citet{esl} a single feature $X_j \in \mathbf{X}$ and its complement set $\mathbf{X}_{\mathcal{P\setminus{j}}} \in \mathbf{X}$ (where $X_j \cup \mathbf{X}_{\mathcal{P\setminus{j}}} = \mathbf{X}$) is considered. $\text{PD}(X_j, g)$ for a given feature $X_j$ is estimated as the average output of the learned function $g(\mathbf{X})$ when all the observations of $X_j$ are set to a constant $x \in \mathcal{X}$ and $\mathbf{X}_{\mathcal{P\setminus{j}}}$ is left unchanged. $\text{ICE}(x_j, \mathbf{x}, g)$ for a given instance $\mathbf{x}$ and feature $x_j$ is estimated as the output of $g(\mathbf{x})$ when $x_j$ is set to a constant $x \in \mathcal{X}$ and all other features $\mathbf{x} \in \mathbf{X}_{\mathcal{P\setminus{j}}}$ are left untouched. PD and ICE curves are usually plotted over some set of constants $x \in \mathcal{X}$, as displayed in Section \ref{sec:res}. Due to known problems for PD in the presence of strong correlation and interactions, PD should not be used alone. PD should always be paired with ICE or be replaced with accumulated local effect (ALE) plots \cite{ice_plots}, \cite{ale_plot}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Discrimination Metrics and Test Description}\label{ssec:di}

@@

Results are laid out for the simulated and mortgage datasets. Accuracy is compared for unconstrained, less interpretable $g^{\text{GBM}}$ and $g^{\text{ANN}}$ models and constrained, more interpretable $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models. Then, for the  $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models, intrinsic interpretabilily, post-hoc explanation, and discrimination testing results are presented.

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/sim_mgbm_loc.png}
\label{fig:sim_mgbm_loc}
\caption{Tree SHAP values for three observations across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for the simulated test data. Tree SHAP values are reported in the margin space, prior to the application of the logit link function.}
\end{figure}

@@

@@

Ridge functions are reminiscent of basis functions ...

\begin{figure}[H]
\centering
\includegraphics[width=17cm]{img/sim_xnn_glob_ridge.png}
\label{fig:sim_xnn_glob_ridge}
\caption{\textbf{A.} Output layer weights, \textbf{B.} corresponding ridge functions, and \textbf{C.} corresponding projection layer weights for $g^\text{XNN}$ on the simulated test data.}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/sim_xnn_loc.png}
\label{fig:sim_xnn_loc}
\caption{Deep SHAP values for three observations across selected percentiles of  $g^\text{XNN}$ on the simulated test data. Deep SHAP values are reported in the probability space, after the application of the logit link function.}
\end{figure}

@@

@@

Results for the mortgage data are presented in Subsections \ref{ssec: c_v_uc_mort} -- \ref{ssec:dis_mort}.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec: c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model performance with respect to unconstrained variants. Assuming that small fit differences do not outweigh the need for intrinsic model interpretability and post-hoc explainability in high-stakes, human-centered, or regulated applications, $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability and post-hoc explainability results are presented along with discrimination testing results. For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of monotonic prediction behavior, and post-hoc Shapley explanation analysis is used to create global and local feature importance. For $g^\text{XNN}$ intrinsic interpretability manifests as plots of sparse weights in the projection and output layers and plots of subnetwork ridge functions. Post-hoc Shapley explanation techniques are also used to generate global and local feature importance for $g^\text{XNN}$. Both $g^\text{MGBM}$ and $g^\text{XNN}$ are evaluated for discrimination using AIR, SMD, and other measures.

\subsubsection{Constrained vs. Unconstrained Model Fit Assessment}\label{ssec: c_v_uc_mort}

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is partially due to the need to present direct comparisons between post-hoc explainability results and the data preprocessing required to use neural networks and tensorflow, e.g. numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced AUC values of $\sim0.81$.

@@

Global feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on LTV ratio, perhaps too high, and also weighs DTI ratio, property value, loan amount, and introductory rate period heavily in many of its predictions.

@@

\noindent The potential over-emphasis of LTV ratio, and the de-emphasis of income, likely an important feature from a business perspective, and the encoded flag features may contribute to the decreased performance of $g^\text{MGBM}$ as compared to $g^\text{XNN}$.

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed on average and across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles using ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the data, and this rigidity could also play a role in the performance differences between $g^\text{MGBM}$ and $g^\text{XNN}$, which does allow for the modeling of non-monotonic interactions. DTI ratio and LTV ratio also appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as the $g^\text{MGBM}$ appears to carry reasonable predictions learned from more populated regions of both features' distributions into the more sparsely populated regions.

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns of ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by the extremely large predictions of high $g^{\text{MGBM}}(\mathbf{X})$ percentile individuals, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$ exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}(\mathbf{X})$ \cite{ice_plots}. Drivers of this interaction can often be elucidated using surrogate decision trees \cite{art_and_sci}. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}(\mathbf{X})$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and in a sparse region of property value. Such fluctuations can be caused by local data characteristics, modeled local interactions, or unfortunately, by overfitting or leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using adverse impact ratio (AIR), marginal effect (ME), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to describe the art of the possible for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

%\setcounter{secnumdepth}{4}

%two part definition
\newcommand{\twopartdef}[4]
{
	\left\{
		\begin{array}{ll}
			#1 & \mbox{if } #2 \\
			#3 & \mbox{if } #4
		\end{array}
	\right.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: AIR, ME, and SMD -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository associated with Sections \ref{sec:m_and_m} and \ref{sec:res} -- \textsection\ref{ssec:soft}
\end{itemize}

To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis.

Post-hoc explanation and discrimination testing techniques are applied to constrained, interpretable models trained on the mortgage data to provide a template workflow for future users of similar methods and tools. Presented explanation techniques include PD, ICE, and Shapley values \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods include AIR, ME, and SMD \cite{feldman2015certifying}, \cite{cohen2013statistical}, \cite{cohen1992power}.\footnote{Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix metrics are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available in the software resources associated delineated in Subsection \ref{ssec:soft}.

@@

Simulated data is created based on a function first proposed in \citet{friedman1979tree} and extended in \citet{friedman1991multivariate}:

\begin{equation}
\begin{aligned}\label{eq:sim}\small
f(\mathbf{X}) = 10\sin(\pi~\mathbf{X}_{\text{Friedman},1}~\mathbf{X}_{\text{Friedman},2}) + 20(\mathbf{X}_{\text{Friedman},3} - 0.5)^2 + 10~\mathbf{X}_{\text{Friedman},4} + 5~\mathbf{X}_{\text{Friedman},5}
\end{aligned}
\end{equation}

\noindent where $\mathbf{X}_{\text{Friedman},j}$ are random uniform features in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output variable for testing spline regression methodologies. In this manuscript, the signal generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be one fifth of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data is generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 rows. The simulated data is then split into a training and test set, with 80,000 and 20,000 observations, respectively.  Within the training set, a 5 fold cross validation indicator is used for training and assessing all models. For an exact specification of the simulated data, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e. the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5 fold cross validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider.  Ultimately, the HMDA data represent the most comprehensive source of data on highly-regulated mortgage lending that is publicly available, which makes it an ideal dataset to use for the types of analyses set forth in Sections \ref{sec:m_and_m} and \ref{sec:res}.

@@

\begin{equation}
\begin{aligned}\label{eq:gbm}
g^{\text{MGBM}}(\mathbf{x}) &= \sum_{b=1}^B T_b\left(\mathbf{x}; \Theta^\text{mono}_b\right)
\end{aligned}
\end{equation}

@@

Following \citet{esl} a single feature $X_j \in \mathbf{X}$ and its complement set $\mathbf{X}_{\mathcal{P} \setminus \{j\}} \in \mathbf{X}$ (where $X_j \cup \mathbf{X}_{\mathcal{P} \setminus \{j\}} = \mathbf{X}$) is considered. $\text{PD}(X_j, g)$ for a given feature $X_j$ is estimated as the average output of the learned function $g(\mathbf{X})$ when all the observations of $X_j$ are set to a constant $x \in \mathcal{X}$ and $\mathbf{X}_{\mathcal{P} \setminus \{j\}}$ is left unchanged. $\text{ICE}(x_j, \mathbf{x}, g)$ for a given instance $\mathbf{x}$ and feature $x_j$ is estimated as the output of $g(\mathbf{x})$ when $x_j$ is set to a constant $x \in \mathcal{X}$ and all other features $\mathbf{x} \in \mathbf{X}_{\mathcal{P} \setminus \{j\}}$ are left untouched. PD and ICE curves are usually plotted over some set of constants $x \in \mathcal{X}$, as displayed in Section \ref{sec:res}. Due to known problems for PD in the presence of strong correlation and interactions, PD should not be used alone. PD should be paired with ICE or be replaced with accumulated local effect (ALE) plots \cite{ice_plots}, \cite{ale_plot}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Discrimination Testing Metrics}\label{ssec:di}

Metrics for testing discrimination in this text are:

\begin{equation}
\label{eq:air}
\begin{aligned}
\text{AIR} \equiv \frac{\text{Pr}(\hat{\mathbf{y}} = 1 | \mathbf{X}_p = 1)}{\text{Pr}(\hat{\mathbf{y}} = 1| \mathbf{X}_c = 1)}
\end{aligned}
\end{equation}

\begin{equation}
\label{eq:me}
\begin{aligned}
\text{ME} \equiv 100 \cdot (\text{Pr}(\hat{\mathbf{y}} = 1| \mathbf{X}_c = 1)  - \text{Pr}(\hat{\mathbf{y}} = 1 | \mathbf{X}_p = 1))
\end{aligned}
\end{equation}

\begin{equation}
\label{eq:smd}
\begin{aligned}
\text{SMD} \equiv \frac{\bar{\hat{\mathbf{y}}}_p - \bar{\hat{\mathbf{y}}}_c}{ \sigma_{\hat{\mathbf{y}}}}
\end{aligned}
\end{equation}

\noindent where $\mathbf{X}_p$ and $\mathbf{X}_c$ represent binary markers created from some demographic attribute $\mathbf{X}_j$, $c$ denotes the control group (often whites or males), $p$ indicates a protected group, and conditional probabilities, $\text{Pr}(\hat{\mathbf{y}}|\mathbf{X}_j = \alpha)$, are evaluated over all $\mathbf{x}^{(i)} \in \{\mathbf{X}_j | x^{(i)}_j = \alpha\}$. Accuracy in each demographic is also considered as a measure of validity for a model and a demographic group.

@@

Results are laid out for the simulated and mortgage datasets. Accuracy is compared for unconstrained, less interpretable $g^{\text{GBM}}$ and $g^{\text{ANN}}$ models and constrained, more interpretable $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models. Then, for the  $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models, intrinsic interpretability, post-hoc explanation, and discrimination testing results are presented.

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/sim_mgbm_loc.png}
\caption{Tree SHAP values for three observations across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for the simulated test data.}
\label{fig:sim_mgbm_loc}
\end{figure}

@@

Ridge functions are reminiscent of basis functions ... distinctive simplistic functions

\begin{figure}[H]
\centering
\includegraphics[width=17cm]{img/sim_xnn_glob_ridge.png}
\label{fig:sim_xnn_glob_ridge}
\caption{\textbf{A.} Output layer $\gamma_k$ weights, \textbf{B.} corresponding $n_k$ ridge functions, and \textbf{C.} associated projection layer $\beta_k$ weights for $g^\text{XNN}$ on the simulated test data.}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/sim_xnn_loc.png}
\caption{Deep SHAP values for three observations across selected percentiles of  $g^\text{XNN}$ on the simulated test data.}
\label{fig:sim_xnn_loc}
\end{figure}

Deep SHAP values are reported in the probability space, after the application of the logit link function.

@@

\begin{table}[H]
\caption{Group size and accuracy for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.}
\centering
\begin{tabular}{cccc}
\toprule
\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} 												& Accuracy \\
\midrule
Protected 1	& 3057.0		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.770 \\ 0.764\end{tabular} \\	
\midrule			 
Control 1		& 16943		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.739 \\ 0.751\end{tabular} \\	
\midrule
Protected 2	& 9916		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.758 \\ 0.761\end{tabular} \\	
\midrule
Control 2		& 10084		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.729 \\ 0.745\end{tabular} \\	
\bottomrule
\label{tab:sim_n_acc}
\end{tabular}
\end{table}

\begin{table}[H]
\caption{AIR, ME and SMD for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.}
\centering
\begin{tabular}{cccccc}
\toprule
\textbf{Model}		& \textbf{Protected Class}								& \textbf{Control Class}							& \textbf{AIR}							& \textbf{ME}								& \textbf{SMD} \\
\midrule
$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} Protected 1 \\ Protected 2\end{tabular}		& \begin{tabular}{@{}c@{}} Control 1 \\ Control 2\end{tabular}	& \begin{tabular}{@{}c@{}} \textcolor{red}{0} \\ \textcolor{red}{0}\end{tabular}	& \begin{tabular}{@{}c@{}} \textcolor{red}{0\%}\\ \textcolor{red}{0\%}\end{tabular}	& \begin{tabular}{@{}c@{}} \textcolor{red}{0} \\ \textcolor{red}{0}\end{tabular}\\
\midrule 
$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} Protected 1 \\ Protected 2\end{tabular}		& \begin{tabular}{@{}c@{}} Control 1 \\ Control 2\end{tabular}	& \begin{tabular}{@{}c@{}} \textcolor{red}{0} \\ \textcolor{red}{0}\end{tabular}	& \begin{tabular}{@{}c@{}} \textcolor{red}{0\%}\\ \textcolor{red}{0\%}\end{tabular}	& \begin{tabular}{@{}c@{}} \textcolor{red}{0} \\ \textcolor{red}{0}\end{tabular} \\													
\bottomrule
\label{tab:sim_disc}
\end{tabular}
\end{table}

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort}.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec: c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model performance with respect to unconstrained variants. Assuming that small fit differences on static test data do not outweigh the need for intrinsic model interpretability and post-hoc explainability in high-stakes, human-centered, or regulated applications, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented. For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. For $g^\text{XNN}$, inherent interpretability manifests as plots of sparse $\gamma_k$ output layer weights, $n_k$ subnetwork ridge functions, and sparse $\beta_j$ weights in the projection layer. Post-hoc Shapley explanation techniques are also used to generate global and local feature importance for $g^\text{XNN}$. Both $g^\text{MGBM}$ and $g^\text{XNN}$ are evaluated for discrimination using AIR, ME, SMD, and other measures.

\subsubsection{Constrained vs. Unconstrained Model Fit Assessment}\label{ssec:c_v_uc_mort}

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g. numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced AUC values of $\sim0.81$.

@@

Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on LTV ratio, perhaps too high, and also weighs DTI ratio, property value, loan amount, and introductory rate period heavily in many of its predictions.

@@

\noindent The potential over-emphasis of LTV ratio, and the de-emphasis of income, likely an important feature from a business perspective, and the encoded no introductory rate period flag feature may contribute to the decreased performance of $g^\text{MGBM}$ as compared to $g^\text{XNN}$.

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the data, and this rigidity could also play a role in the performance differences between $g^\text{MGBM}$ and $g^\text{XNN}$, which does allow for the modeling of non-monotonic interactions. DTI ratio and LTV ratio also appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$, and lower,  exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and in a sparse region of property value. Such fluctuations can be caused by local data characteristics, modeled local interactions, or unfortunately, by overfitting or leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features. \textcolor{red}{This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}(\mathbf{X})$ \cite{ice_plots}, and drivers of such interactions can often be elucidated using surrogate decision trees \cite{art_and_sci}.
}

In Figure \ref{fig:mort_mgbm_loc}, local Tree SHAP values are displayed for selected individuals at the 10\textsuperscript{th}, 50\textsuperscript{th}, and 90\textsuperscript{th} percentiles of $g^{\text{MGBM}}(\mathbf{X})$ in the mortgage test data. The selected individuals show an expected progression of mostly negative Shapley values at the 10\textsuperscript{th} percentile, a mixture of positive and negative Shapley values at the 50\textsuperscript{th} percentile, mostly positive Shapley values the 90\textsuperscript{th} percentile, and with globally important features driving most local model decisions.
